---
title: "GammaSleep Supplementary Analysis"
author: "L. Hainke, J. Dowsett, M. Spitschan, J. Priller"
date: '`r Sys.Date()`'
output:
  pdf_document: default
  html_document:
    code_folding: hide
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, fig.height=4.5, fig.width=7, fig.align="left")
```

```{r Environment Setup, warning=FALSE, message=FALSE}
# Working directory
setwd("C:\\Users\\Mitarbeiter\\Documents\\Gamma_Sleep\\Github_Repo\\Gamma-Sleep\\Code\\Statistics")

# Function files in current directory
source("GammaSleep_statistics_functions.r")
source("GammaSleep_plots_functions.r")
source("GammaSleep_data-handling_functions.r")

# Libraries
library(chron)
library(ggplot2)


## Global variables
# Using global environment (<<- assignment) to avoid redundant function arguments

# Path to folders with derivative data
path_derivatives <<- "C:\\Users\\Mitarbeiter\\Documents\\Gamma_Sleep\\Data\\Derivatives_supplementary\\" 

# List of folder names in directory, corresponding to participant numbers
list_IDs <<- list.dirs(path = path_derivatives, full.names = FALSE, recursive = FALSE)

# Remove rejected datasets
list_IDs <<- list_IDs[! list_IDs %in% c("03","15")] # subjects 03 and 15 were drop-outs
```

```{r Dataframes for Derivative Data}
## Initializing 
data_demo <<- initialize_dataframe("demographic")

data_sleep <<- initialize_dataframe("sleep_quality")

data_PSD <<- initialize_dataframe("PSD_metrics")
                        
data_SSVEP <<- initialize_dataframe("SSVEP_metrics")

## Loading data
load_derivative_data()
```

# Rationale

Linear interpolation refers to a data cleaning step employed to remove periodic artifacts. Here, such artifacts have been caused by the LEDs of the custom sleep mask placed directly over subjects' eyes, and therefore close to EEG electrodes. At the intended stimulation frequency of 40 Hz and a duty cycle of 50 %, the LEDs were programmed to go ON at timepoint 0 of each 25 ms segment, and OFF after 12.5 ms. The mask was built to be shielded as well as possible, however, SSVEPs were computed from a very high number of segments. While this is favourable to average out noise overall, any electric contamination by the LEDs would be very consistent in timing, and therefore visible in the averaged time series.  
The image below shows an SSVEP particularly affected by this issue, with one downward and one upward peak distorting the underlying sinusiodal waveform. The red dashed lines simbolize the linear interpolation procedure: Essentially, data at the beginning (timepoints [-1;3]) and middle (timepoints [11;15]) of each segment is replaced with a straight line. Averaging across "cleaned" segments then results in a sinusoidal form, without residuals of LED ON/OFF artifacts.  

```{r Plot linear interpolation, warning=FALSE}
# Load data
SSVEP = read.csv("C:\\Users\\Mitarbeiter\\Documents\\Gamma_Sleep\\Data\\Derivatives_main\\24\\Experimental\\24_experimental_SSVEP-output-curves.csv")
SSVEP = SSVEP[,c(1,3)] # keep time points & SSVEP for N2_exp (most affected)

# Linear interpolation time points
interpolation_timepoints = c(-1,3,11,15)

# Plot
ggplot(SSVEP, aes(x=X, y=N2_SSVEP)) +
  geom_line(linewidth=1) + # plot SSVEP
  geom_segment(x=interpolation_timepoints[1], y=-0.09, xend=interpolation_timepoints[2], yend=SSVEP$N2_SSVEP[SSVEP$X==interpolation_timepoints[2]], colour="red", linetype="dashed", size=1.2) + # First interpolation segment
  geom_segment(x=interpolation_timepoints[3], y=SSVEP$N2_SSVEP[SSVEP$X==interpolation_timepoints[3]], xend=interpolation_timepoints[4], yend=SSVEP$N2_SSVEP[SSVEP$X==interpolation_timepoints[4]], colour="red", linetype="dashed", size=1.2) + # Second interpolation segment
  labs(title="Example SSVEP affected by artifacts", y="Amplitude (uV)", x="Time (ms)") +
  scale_x_continuous(limits=c(-1,24), breaks=c(0,5,10,15,20,25)) +
  ylim(-0.12,0.12) +
  theme_minimal(base_size=7) +
  theme(panel.grid.minor = element_blank(), # remove background lines
        strip.text.x = element_text(size=rel(3), face="bold"),
        axis.title.x = element_text(size=rel(2), vjust=-1), # Label sizes, relative
        axis.title.y = element_text(size=rel(2), vjust=1.5),
        axis.text = element_text(size=rel(1.5)), 
        plot.title = element_text(size=rel(3), vjust=1.5),
        plot.margin = margin(0.5,0.5,0.5,0.5, "cm"))

```

This procedure was not defined in the preregistration, since pilot data were not affected by artifacts. In fact, neither was the majority of the datasets. Given that the presence or absence of artifacts is constant within participants, and comparisons between conditions were carried out within-subjects, artifacts do not distort pairwise comparison results. For those reasons, linear interpolation was not applied to main analyses. Nevertheless, since some SSVEPs showed these characteristic peaks, we report results with this processing step applied here.  
In short, **the results of confirmatory analyses do not change**.  

# Confirmatory Analysis [EEG Power Spectral Density]

## H1: W_exp > W_con

During wakefulness, 40 Hz power was expected to be higher in the experimental condition (session 01) than in the control condition (session 02). This hypothesis was **confirmed**, with a **large effect** size:  

```{r H1}
test_two_levels(data_PSD$W_PSD40_exp, data_PSD$W_PSD40_con, parametric=TRUE, alternative="greater")
```

## H2: N2_exp > N2_con

During sleep stage N2, 40 Hz power was expected to be higher in the experimental condition (session 03) than in the control condition (session 02). This hypothesis was **confirmed**, with a **medium effect** size:  

```{r H2}
test_two_levels(data_PSD$N2_PSD40_exp, data_PSD$N2_PSD40_con, parametric=TRUE, alternative="greater")
```

## H3: N3_exp > N3_con

During sleep stage N3, 40 Hz power was expected to be higher in the experimental condition (session 03) than in the control condition (session 02). This hypothesis was **confirmed**, with a **medium effect** size:  

```{r H3}
test_two_levels(data_PSD$N3_PSD40_exp, data_PSD$N3_PSD40_con, parametric=TRUE, alternative="greater")
```

## H4: REM_exp > REM_con

During sleep stage REM, 40 Hz power was expected to be higher in the experimental condition (session 03) than in the control condition (session 02). This hypothesis was **confirmed**, with a **large effect** size:  

```{r REM}
test_two_levels(data_PSD$REM_PSD40_exp, data_PSD$REM_PSD40_con, parametric=FALSE, alternative="greater")
```

## H5: W_exp =/= N2_exp =/= N3_exp =/= REM_exp

In the experimental condition, 40 Hz power was expected to differ between stages. This hypothesis was **confirmed**, with a **large effect** size. However, post-hoc tests reveal that **only stage W differs** from all other stages, i.e., the stimulation effect is largest in W. There is no significant difference between the sleep stages N2, N3, and REM.

```{r H5}
# Conversion to long format
data_PSD_dB_long = dataframe_wide_to_long(data_PSD[,c("ID","W_PSD40_con","N2_PSD40_con","N3_PSD40_con","REM_PSD40_con","W_PSD40_exp","N2_PSD40_exp","N3_PSD40_exp","REM_PSD40_exp")])

# Get data subset
data_anova_PSD_dB = subset(data_PSD_dB_long, condition=="exp", select=c(ID,stage,value))

# Run statistical test
test_multiple_levels(data_anova_PSD_dB, parametric=TRUE, sphericity_correction="greenhouse-geisser")
```

## Signal-to-Noise Ratio

Here are the **median SNR values**. Consistent with confirmatory analyses, all SNR values in the control condition are ~1, and all SNR values in the experimental condition are >1, with a markedly higher value in W~exp~. 

```{r SNR PSD medians}
# Conversion to long format
data_PSD_SNR_long = dataframe_wide_to_long(data_PSD[,c("ID","W_SNR40_con","N2_SNR40_con","N3_SNR40_con","REM_SNR40_con","W_SNR40_exp","N2_SNR40_exp","N3_SNR40_exp","REM_SNR40_exp")])

# Median SNR values per condition
medians_PSD_SNR = aggregate(value ~ stage * condition, data_PSD_SNR_long, median) # mean likely biased by outliers
print(medians_PSD_SNR, row.names=FALSE)
```

# SSVEPs

The following plots depict the **SSVEPs, interpolated and averaged** across all subjects, by stage and condition.

```{r SSVEP avg}
# All SSVEP time series data in long format
timeseries_SSVEPs = import_SSVEPs_to_long_dataframe()

# Collapse data over subjects
data_SSVEPs_avg = timeseries_SSVEPs %>% group_by(time,stage,condition) %>% summarise_at(vars("value"), mean)

# Separate stage W
data_SSVEPs_avg_wake = data_SSVEPs_avg[data_SSVEPs_avg$stage=="W",]
data_SSVEPs_avg_sleep = data_SSVEPs_avg[data_SSVEPs_avg$stage!="W",]
```
  
```{r SSVEP avg W}
plot_avg_SSVEPs_stage(data_SSVEPs_avg_wake, title_plot="Average SSVEPs - Wake")
```
  
```{r SSVEP avg Sleep}
plot_avg_SSVEPs_stage(data_SSVEPs_avg_sleep, title_plot="Average SSVEPs - Sleep")
```

Here are the **median SNR values**. Consistent with confirmatory analyses, all SNR values in the control condition are ~1, and all SNR values in the experimental condition are >1, with a markedly higher value in W_exp. 

```{r SNR SSVEP medians}
# Conversion to long format
data_SSVEP_SNR_long = dataframe_wide_to_long(data_SSVEP[,c("ID","W_SNR_con","N2_SNR_con","N3_SNR_con","REM_SNR_con","W_SNR_exp","N2_SNR_exp","N3_SNR_exp","REM_SNR_exp")])

# Median SNR values per condition
medians_SSVEP_SNR = aggregate(value ~ stage * condition, data_SSVEP_SNR_long, median) # mean likely biased by outliers
print(medians_SSVEP_SNR, row.names=FALSE)
```





